---
title: "MENTAL HEALTHCARE"
subtitle: "ASSIGNMENT 11.3"
author: "Siddhartha Bhaumik"
date: '2022-06-04'
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

-   Mental health disorders are one of the leading health issues in the United States and 
    it affects almost 10% of the population. This has significantly increased since the beginning of Covid19 pandemic 
    and is not just seen in adults but children and young adults as well.

-   Mental health awareness is another significant problem as lot of people are unaware of their problems until 
    those turn into severe health issues like anxiety, depression, and other life-threatening risks.

-   Also, many people don’t know how to or from where to get help. Or they feel ashamed talking about their 
    condition as that may impact their relationship both personally and professionally as well as their growth 
    in life and at workplace.

-   Another important thing to note is that in U.S there is a shortage in certified mental health professionals.

-   Data Science/Artificial Intelligence can play an important part here by bridging some of the current gaps 
    in Mental Healthcare sector.
    
## Problem Statement

-   Mental health is often overlooked which many times leads up to serious health issues. 
    This is not specific to any age group as all are vulnerable. Based on the statistics, around 10% of U.S population has     some form of mental health issue out of which up to 20% are children and young adults. 
    
-   Awareness and acceptance is the key to address this problem.    

## Steps taken To Address This Problem

-   Easy access to counseling and proper guidance at every stage of life. Be it at home by parents, at school by         teachers/professionals, at workplace by professionals, etc.
-   Bring awareness and acceptance so that people who need help or people around you can identify this at an early stage and find support from friends/family and professionals. This can be done through wellness programs and seminars, etc.
-   Majority of people have smartphones. AI mobile apps can be a great solution for self-assessment as well as therapy and other needs.
-   Several fitness gadgets are in market which can track your sleep, heart rate, blood pressure, etc. and can share that information through apps which can further evaluate and predict your overall health.
-   With so much patient data available digitally like health reports, lab reports, social media interactions, etc., different AI tools and techniques can analyze patient’s data and flag physical and mental states. This can help in early detection and remedies.

## Analysis
```{r echo=TRUE}

## Load required package
library(ggplot2)
library(pastecs)
library(dplyr)
library(purrr)
library(stringr)
library(factoextra)  # clustering algorithms & visualization
library(cluster)  # clustering algorithms
theme_set(theme_minimal())


## Load 'Any Mental Illness in the Past Year Data' from 51 US states to
state_any_mental_df <- read.csv("/Users/siddharthabhaumik/Documents/GitHub/dsc520/map_data.csv")


## Load 'Major Depressive Episode in the Past Year Data' from 51 US states to
state_dep_mental_df <- read.csv("/Users/siddharthabhaumik/Documents/GitHub/dsc520/data/map_data_rcrd.csv")


## Load 'Received Mentalhealth services in Past Year Data' from 51 US states to
state_rcvd_mental_df <- read.csv("/Users/siddharthabhaumik/Documents/GitHub/dsc520/data/map_data_dep.csv")

### Covid19 US dataset
covid19_us_df <- readxl::read_excel("/Users/siddharthabhaumik/Documents/GitHub/dsc520/data/search_term_us.xlsx")

## Load the 'Mental Health in Tech Survey' to
tech_survey_df <- read.csv("/Users/siddharthabhaumik/Documents/GitHub/dsc520/survey.csv")

## Viewing US Sample data related to 'Covid19 & Mental health effect/awareness'
head(covid19_us_df)

## summary
summary(covid19_us_df)

## plot
boxplot(covid19_us_df[-1], col = "red", varwidth = TRUE, xlab = "Mental Health Disorders")

# During Covid19 pandemic, The mental health related search terms like "depression", 
# "anxiety", "ocd", etc. increased significantly on google search engine.

# This shows people either impacted directly or hearing from someone from their family
# and friends. But, it shows increased awareness and looking for options on the internet 
# to address this problem.
# From this dataset I can see Depression, Anxiety and counseling are the top 3 search 
# items but other than that not much information can be extracted from this dataset.

### US States Mental Heath Data

## Viewing Sample data from 'Any mental illness/past year' dataset
head(state_any_mental_df)

str(state_any_mental_df)

## Plot
ggplot(state_any_mental_df, aes(x = state, y = estimate)) +
  geom_point(size = 2.5, color = "red") + 
  geom_segment(aes(x = state, xend = state, y = 0, yend = estimate)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.7)) # Rotate axis label

# The top 3 states with any mental illness in year 2018-19 are Utah, West Virginia & Oregon.

## Viewing Sample data from 'Major depressive episode/past year' dataset
head(state_dep_mental_df)

## Plot
ggplot(state_dep_mental_df, aes(x = state, y = estimate)) +
  geom_point(size = 2.5, color = "red") + 
  geom_segment(aes(x = state, xend = state, y = 0, yend = estimate)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.7)) # Rotate axis label

#The top 3 states with major depressive episode in year 2018-19 are Utah, Oregon & West Virginia.

## Viewing Sample data from 'Received Mental health services/past year' dataset
head(state_rcvd_mental_df)

#Plot
ggplot(state_rcvd_mental_df, aes(x = state, y = estimate)) +
  geom_point(size = 2.5, color = "red") + 
  geom_segment(aes(x = state, xend = state, y = 0, yend = estimate)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.7)) # Rotate axis label

#The top 3 states with people receiving mental health related services in year 2018-19 
# are Vermont, Massachusetts & Maine.

# So, based on the plots, we can see that Utah, Oregon and West Virginia are the 
# top 3 states for any mental health illness as well as major depressive episodes.

# But, from treatment perspective, none of the 3 are in the top 3 list. 
# This shows that even though lot of cases are reported, still not everyone is 
# availing any kind of treatments.

### Mental health at Work place

tech_survey_upd_df <- tech_survey_df %>% filter(Country == "United States") %>% filter(Age > 12) %>% select(Age,Gender,family_history,treatment,remote_work,work_interfere,benefits,wellness_program,seek_help,anonymity,mental_health_consequence,obs_consequence)

head(tech_survey_upd_df)

# Standardize Gender with Male, Female, Other
tech_survey_upd_df["Gender"][tech_survey_upd_df["Gender"] == "M" | tech_survey_upd_df["Gender"] == "m" 
                             | tech_survey_upd_df["Gender"] == "male" | tech_survey_upd_df["Gender"] == "mail" | tech_survey_upd_df["Gender"] == "Mail"
                             | tech_survey_upd_df["Gender"] == "Cis male" | tech_survey_upd_df["Gender"] == "Cis Male"
                             | tech_survey_upd_df["Gender"] == "Male-ish" | tech_survey_upd_df["Gender"] == "Male (CIS)"
                             | tech_survey_upd_df["Gender"] == "Man" | tech_survey_upd_df["Gender"] == "Cis Man"
                             | tech_survey_upd_df["Gender"] == "Malr" | tech_survey_upd_df["Gender"] == "msle"
                             | tech_survey_upd_df["Gender"] == "Mal" | tech_survey_upd_df["Gender"] == "Make"
                             | tech_survey_upd_df["Gender"] == "maile" | tech_survey_upd_df["Gender"] == "cis male"] <- "Male"


tech_survey_upd_df["Gender"][tech_survey_upd_df["Gender"] == "F" | tech_survey_upd_df["Gender"] == "f" 
                             | tech_survey_upd_df["Gender"] == "female" | tech_survey_upd_df["Gender"] == "femail"
                             | tech_survey_upd_df["Gender"] == "Cis Female" | tech_survey_upd_df["Gender"] == "cis-female/femme"
                             | tech_survey_upd_df["Gender"] == "Woman" | tech_survey_upd_df["Gender"] == "Female (cis)"
                             | tech_survey_upd_df["Gender"] == "Femake" | tech_survey_upd_df["Gender"] == "woman" 
                             | tech_survey_upd_df["Gender"] == "Female (trans)"] <- "Female"

tech_survey_upd_df["Gender"][tech_survey_upd_df["Gender"] == "Female (trans)" | tech_survey_upd_df["Gender"] == "queer/she/they" 
                             | tech_survey_upd_df["Gender"] == "non-binary" | tech_survey_upd_df["Gender"] == "Nah"
                             | tech_survey_upd_df["Gender"] == "Genderqueer" | tech_survey_upd_df["Gender"] == "Trans-female"
                             | tech_survey_upd_df["Gender"] == "Trans woman" ] <- "Others"

head(tech_survey_upd_df)


# Only considering survey results from United States as its the majority.
# Noticed some negative numbers under 'Age' column which I will filter out.
# Under 'Gender' column, I see lot of variation and spelling error like Male,Mail,maile,
# M, Cis Male, Female, Cis Female, etc. I will make it consistent as Male, Female, Other.  
# Dropped some columns like State, No of Employee, Tech company, etc. as I don't think 
# they add much value.  

# Basically I am looking for how many people opted for 'Treatment'.

## Viewing Mental health in US Tech industry
head(tech_survey_upd_df)

#summary
summary(tech_survey_upd_df)

str(tech_survey_upd_df)

# Replace Gender, Wellness program & Seek help column values to numeric
tech_survey_upd_df["Gender"][tech_survey_upd_df["Gender"] == "Male "] <- 1

tech_survey_upd_df["Gender"][tech_survey_upd_df["Gender"] == "Female "] <- 2

tech_survey_upd_df["Gender"][tech_survey_upd_df["Gender"] == "Others"] <- 3

tech_survey_upd_df["family_history"][tech_survey_upd_df["family_history"] == "Yes"] <- 1

tech_survey_upd_df["family_history"][tech_survey_upd_df["family_history"] == "No"] <- 2

tech_survey_upd_df["wellness_program"][tech_survey_upd_df["wellness_program"] == "Yes"] <- 1
tech_survey_upd_df["wellness_program"][tech_survey_upd_df["wellness_program"] == "No"] <- 2
tech_survey_upd_df["wellness_program"][tech_survey_upd_df["wellness_program"] == "Don't know"] <- 3

tech_survey_upd_df["seek_help"][tech_survey_upd_df["seek_help"] == "Yes"] <- 1
tech_survey_upd_df["seek_help"][tech_survey_upd_df["seek_help"] == "No"] <- 2
tech_survey_upd_df["seek_help"][tech_survey_upd_df["seek_help"] == "Don't know"] <- 3


# Filter out age above 100
tech_survey_upd_df2 <- tech_survey_upd_df %>% filter(Age < 100) %>%  select(Age,Gender,family_history,treatment,wellness_program,seek_help)

tech_survey_upd_df2["treatment"][tech_survey_upd_df2["treatment"] == "Yes"] <- 1
tech_survey_upd_df2["treatment"][tech_survey_upd_df2["treatment"] == "No"] <- 2

# Plot
ggplot(tech_survey_upd_df2, aes(x=treatment, y=Age, col=Gender)) + geom_point() + ggtitle("Treatment vs. Age vs. Gender") + xlab("Treatment (Yes = 1, No = 2)") + ylab("Age")

ggplot(tech_survey_upd_df2, aes(x=treatment, y=family_history, col=wellness_program)) + geom_point() + ggtitle("Treatment vs. family_history vs. wellness_program") + xlab("Treatment (Yes = 1, No = 2)") + ylab("family_history (Yes = 1, No = 2)")

ggplot(tech_survey_upd_df2, aes(x=treatment, y=wellness_program)) + geom_point() + ggtitle("Treatment vs. wellness_program") + xlab("Treatment (Yes = 1, No = 2)") + ylab("wellness_program (Yes = 1, No = 2, Don't Know = 3)")

ggplot(tech_survey_upd_df2, aes(x=treatment, y=family_history, col=seek_help)) + geom_point() + ggtitle("Treatment vs. family_history vs. Seek_help") + xlab("Treatment (Yes = 1, No = 2)") + ylab("family_history (Yes = 1, No = 2)")

plot(tech_survey_upd_df2)

# My target variable here is 'treatment'.Workplaces which promotes mental health and 
# support their employees through different benefits and wellness programs will see more
# people opting for treatments and other kind of help if needed.

# I see more Male population tied to mental health but I see the ratio of male population 
# much higher than female and others in the dataset.
# This can be due to more Male working in the tech industry or the survey population is biased.

# I also looked at family history to see if those people tend to be more aware qnd 
# seek help if needed but didnt see much correlation.
```

## Limitations

- The survey datasets picked up for Covid19 doesn't have much details to link it with other US State mental health datasets. Also, I was looking for overall workplace dataset and not specifically tech industry dataset.

## Concluding Remarks    

-   AI is going to revolutionize the mental health care system due to its accessibility, affordability, efficiency, accuracy, and support. Despite some of the drawbacks related to overall data privacy, I feel AI is the key to control this problem through smartphones and other smart devices accessible to people and wellness/healthcare professionals in near future.
    

## Datasets/Citations

-   "COVID-19 and Mental Health Search Terms" dataset from Kaggle. <https://www.kaggle.com/datasets/luckybro/mental-health-search-term>     The search interest of mental health related terms on Google before and after the outbreak of COVID-19 pandemic\
    reveals how public's concern is affected by the pandemic, and its impact to mental health of people around the world.

-   "Mental Health in Tech Survey" dataset from Kaggle <https://www.kaggle.com/datasets/osmi/mental-health-in-tech-survey> 
    This dataset is from a 2014 survey that measures attitudes towards mental health and frequency of mental health\
    disorders in the tech workplace.

-   "Any Mental Illness in the Past Year among Adults Aged 18 or Older, by State: 2018-2019" dataset from SAMHDA.gov <https://pdas.samhsa.gov/saes/state> 
    This dataset is maintained by 'Substance Abuse & Mental Health Data Archive' government agency and contains any type of 
    mental health related issues in adults aged 18 and older for the year 2018-2019.



